---
title: "R practicals"
author: "Malay K Basu"
date: "11/8/2020"
output: html_notebook:
         toc:true
editor_options: 
  chunk_output_type: inline
---
# R data structures

Assign the value of `3` to the variable `x`.
```{r}
x <- 3
x
```

Let’s create another variable called y and give it a value of 5.
```{r}

```

What do you get in the console for the following operation:
```{r}
x + y
```

Try assigning the results of this operation to another variable called number.
```{r}
number <- x + y
```

Try changing the value of the variable `x` to 5. What happens to `number`?
```{r}

```

Now try changing the value of variable `y` to contain the value 10. What do you need to do, to update the variable `number`?
```{r}

```

Let’s create a vector of genome lengths and assign it to a variable called `glengths`.
```{r}
glengths <- c(4.6, 3000, 50000)
glengths
```

Create a character vector and store the vector as a variable called `species`.
```{r}
species <- c("ecoli", "human", "corn")
species
```

Try to create a vector of numeric and character values by combining the two vectors that we just created (`glengths` and `species`). Assign this combined vector to a new variable called `combined`. Hint: you will need to use the combine `c()` function to do this. Print the `combined` vector in the console, what looks different compared to the original vectors?
```{r}

```

  
Let's say that in our experimental analyses, we are working with three different sets of cells: normal, cells knocked out for geneA (a very exciting gene), and cells overexpressing geneA. We have three replicates for each celltype.

 1. Create a vector named `samplegroup` with nine elements: 3 control ("CTL") values, 3 knock-out ("KO") values, and 3 over-expressing ("OE") values.
 2. Turn `samplegroup` into a factor data structure.
```{r}

```

Create a data frame called `favorite_books` with the following vectors as columns:
```{r} 
titles <- c("Catch-22", "Pride and Prejudice", "Nineteen Eighty Four")
pages <- c(453, 432, 328)
```

Create a list called `list2` containing `species`, `glengths`, and `number`.
```{r}

```


# Functions and arguments

Let's use base R function to calculate **mean** value of the `glengths` vector. You might need to search online to find what function can perform this task.
```{r}

```

Create a new vector.
```{r}
test <- c(1, NA, 2, 3, NA, 4)
```
Use the same base R function from last excersize calculate mean value of the `test` vector. The output should be `2.5`. Deal with `NA` values properly.
```{r}

```

Another commonly used base function is `sort()`. Use this function to sort the `glengths` vector in **descending** order.
```{r}

```

Write a function called `multiply_it`, which takes two inputs: a numeric value `x`, and a numeric value `y`. The function will return the product of these two numeric values, which is `x * y`. For example, `multiply_it(x=4, y=6)` will return output `24`.
```{r}

```

Re-write the function `multiply_it`, where `y` is set with a **default** value of 6 (hint: assign the value when defining the input). Apply this function by providing only one input `x`. For example, `multiply_it(x=4)` will return output `24`. Also test if you can **overwrite** the default value of `y` by providing both `x` and `y` values as arguments in the function. For example, `multiply_it(x=4, y=8)` will return output `32`, but not `24`.
```{r}

```

# Reading in and inspecting data

Download this tab-delimited .txt file https://github.com/cb2edu/CB2-101-BioComp/raw/2020/02-Introduction_to_R/iris.txt) and save it in your project’s data folder.
```{r}
download.file("https://github.com/cb2edu/CB2-101-BioComp/raw/2020/02-Introduction_to_R/iris.txt",destfile = "iris.txt")
```

Read it in to R using read.table() and store it as the variable `iris`, keeping in mind that: 
 
 1. all the columns have column names 
 2. Display the contents of proj_summary in your console
```{r}

```

Use the class() function on `Species` and `Petal.length`, how does the output differ between the two?
```{r}

```

Use the summary() function on the `iris` dataframe, what is median length for each of the `numeric` column?
```{r}

```

How long is the Species factor?
```{r}

```

What are the dimensions of the iris dataframe?
```{r}

```

How many elements in (how long is) the output of colnames(iris)? Don’t count, but use another function to determine this.
```{r}

```

# Data wrangling

Extract only those elements  `Sepal.Length` that are not `setosa`.
```{r}

```

Return the `Sepal.Length` and `Petal.Length` columns for `setosa` and `versicolor`.
```{r}

```

Return the fourth and ninth values of the `Petal.Length` column of the `iris` dataframe.
```{r}

```

Extract the `Petal.Length`  column as a data frame.
```{r}

```

Subset the `iris` dataframe to return only the rows of data with a species is `setosa`.
```{r}

```


Create a list named `random` with the each element being the `Petal.Length` for each species of the `iris` dataframe.
```{r}

```

Extract the first component from the list.
```{r}

```

Set names for the random list you created in the last exercise to each species.
```{r}

```

Extract the `Petal.Length` using the $ notation for each species.
```{r}

```

# Reordering

Let’s create the vectors first and second.
```{r}
first <- c("A","B","C","D","E")
second <- c("B","D","E","A","C")  # same letters but different order
```

How would you reorder the second vector to match first?
```{r}

```

# Generating a Boxplot with ggplot2

Covert the `iris` data frame to be used in `ggplot2`.
```{r}
library(ggplot2)
library(tidyr)
data(iris)
new_iris <- data.frame(species=iris$Species, petal.length=iris$Petal.Length)
View(new_iris)
```

Generate a boxplot using the data in the new dataframe.
```{r}

```
Add a title to the plot and center the title.
```{r}

```

Rewrite the axis labels in publishable format format (increase the font size and use proper capitalization). 
```{r}

```

Change the default theme to black and white.
```{r}

```



# Writing to file and exporting plots

Use the `ggsave` function to write the plot to a PDF file.
```{r}

```

Write the `new_iris` data set to a `.tsv` file.
```{r}

```




