---
title: "Singular Value Decomposition Demo"
author: "Malay Basu (malaykbasu@gmail.com"
date: "2023-11-20"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This is a demostration of singular value decomposition. Check the talk for the theretical details of the SVD. This demo just shows how the SVD can be used for image compression. 

We need to install imager library to open an image.

```{r}
ifelse(require(imager), suppressPackageStartupMessages(library(imager)), BiocManager::install("imager"))
```

Now we will download an image file for our use.

```{r}
download.file(url = "https://upload.wikimedia.org/wikipedia/commons/6/6f/Editorial_cartoon_depicting_Charles_Darwin_as_an_ape_%281871%29.jpg", destfile = "darwin.jpg", quiet = T)
```

Here is the image that we will use.

```{r}
darwin <- load.image("darwin.jpg")
plot(darwin, axes = F)
```



To original image is a mixture of 3 channels. To find a proper variance for the overall image, we will convert it to grayscale first.

```{r}
darwin_gray <- grayscale(darwin)
plot(darwin_gray, axes=F)
```

Check the dimesion of this image.

```{r}
dim(darwin_gray)
```

Now we'll save the grascale image as PNG and check the size.

```{r}
save.image(darwin_gray, "darwin_gray_before_svd.png")

```

Check the file size.

```{r}
file.info("darwin_gray_before_svd.png")$size/1024^2
```



SVD will decompse the image into 3 matrices. The middle diagnoal matrix holds the variance information. We will do an SVD of the image and plot a scree plot.

```{r}
svd1 <- svd(scale(darwin_gray))
str(svd1)
```

d is the diagolar vector. Let's plot the scree plot.

```{r}
plot( svd1$d^2/sum(svd1$d^2), xlab= "Igen vector", ylab="Variance explained")
```

Let's plot using various number of SVD.

```{r}
par(mfrow=c(2,3))

for (i in c(5, 10, 25, 50, 100, length(svd1$d))) {
    reconstruted <- svd1$u[, 1:i] %*% diag(svd1$d[1:i]) %*% t(svd1$v[, 1:i])
    plot(as.cimg(reconstruted), axes=F, main=paste("Eigenvectors", i, sep=" "))
}

```

We can see that approximately 50 eigenvectors can reconstructued the full image.

```{r}
compressed <- svd1$u[, 1:50] %*% diag(svd1$d[1:50]) %*% t(svd1$v[, 1:50])
plot(as.cimg(compressed), axes=F)
```


```{r}
dim(compressed)
```

Save the file and check the size.

```{r}
save.image(as.cimg(compressed), "darwin_gray_after_svd.png")
```

Size of the file.
```{r}
file.info("darwin_gray_after_svd.png")$size/1024^2

```

